---
title: "Basics: Working With Data in R"
code-fold: false
---

# Introduction

This page covers the most practical basics you need before doing any statistics
in R:

- loading data
- exploring structure
- handling missing values
- descriptive statistics
- basic tables and proportions
- random number generation
- checking simple distribution shapes

You can use any dataset, but for this tutorial we work with:

**`perfectionism_total_scores.csv`**
Small psychological dataset with one continuous score variable.

<p>
  <a href="../assets/data/perfectionism_total_scores.csv" download>
    Download the dataset (perfectionism_total_scores.csv)
  </a>
</p>

```{r setup}
library(tidyverse)
df <- read_csv2("../assets/data/perfectionism_total_scores.csv")
```

## 1. First look at the data

Understanding your dataset is the very first step. Look at the variable names
and data types. What type is the main score variable?

```{r explore}
glimpse(df)
summary(df)
head(df)
```


## 2. Descriptive statistics

Compute the mean, median, and standard deviation of the score.

```{r descriptives}

# Mean
mean(df$FMPS_total, na.rm = TRUE)

# Median
median(df$FMPS_total, na.rm = TRUE)

# Standard deviation
sd(df$FMPS_total, na.rm = TRUE)

# Range
range(df$FMPS_total, na.rm = TRUE)

# Quantile
quantile(df$FMPS_total, probs = c(.25, .5, .75), na.rm = TRUE)

# Or all in one table
df %>%
    select(FMPS_total:DASS_D_Av) %>%
    psych::describe() %>%
    rownames_to_column("Scale") %>%
    as_tibble()

```

## 3. Missing values (NA)

Remove NA values from score and compute the mean of the cleaned vector.

```{r missing_values}
sum(is.na(df$FMPS_total))
df_noNA <- df %>% drop_na(FMPS_total)
nrow(df) - nrow(df_noNA)
```

## 4. Frequency tables & proportions

For categorical variables we would use table() and prop.table().
Here we artificially convert scores into categories:

Convert scores into three groups (your own cut points) and compute their
proportions.

```{r frequency_tables}
df_cat <- df %>%
  mutate(level = case_when(
    FMPS_total < 30 ~ "Low",
    FMPS_total < 60 ~ "Medium",
    TRUE ~ "High"
  ))

table(df_cat$level)
prop.table(table(df_cat$level)) %>% round(3)
```

## 5. Random numbers & set.seed()

Generate 20 random values from a normal distribution with mean 70 and sd 15.

```{r random_numbers}
set.seed(42)
x <- rnorm(100, mean = 50, sd = 10)
head(x)
```

Why set.seed()?
It makes randomness reproducible.

## 6. Quick normality check

Not a hypothesis test — just a visual check.

Inspect the histogram and Q-Q plot. Does the distribution look roughly normal?

```{r qqplot}
df %>%
  ggplot(aes(x = FMPS_total)) +
  geom_histogram(bins = 20, fill = "#5c6bc0") +
  theme_minimal()

df %>%
  ggplot(aes(sample = FMPS_total)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()
```

## Take-home message

- Always start with glimpse(), summary(), head()
- Use na.rm = TRUE when needed
- table() + prop.table() are essential for categorical summaries
- set.seed() ensures reproducible randomness
- Quick plots often reveal more than formal tests