
---
title: "Correlation in R"
code-fold: false
---

This page walks through several ways of working with correlations in R,
using two small teaching datasets:

- an **ice-cream experiment** (puzzle vs. video game scores)
- a **US crime dataset** (crime rate, demographics, wages, etc.)

The goal is not only to compute a single `cor()` value, but to learn
how to:

- compute Pearson, Spearman, and Kendall correlations,
- test correlations with `cor.test()`,
- compute correlations **separately by groups**,
- explore larger sets of variables with **correlation matrices** and
  helper functions.

## Setup

```{r}
library(tidyverse)


# install.packages("corrplot")
# install.packages("lares")
library(corrplot)
library(lares)

# Load datasets
# Adjust paths to match your project structure
icecream <- read_csv("../assets/data/ice_cream.csv")
crime    <- read_csv("../assets/data/crime.csv")
```

---

## 1. Ice-cream experiment: basic correlations

In this dataset, participants completed:

- a **puzzle task** (`puzzle`), and  
- a **video game task** (`video`).  

We start with the simplest question:

> How strongly are puzzle scores and video game scores related?

### 1.1 Pearson correlation

```{r}
cor(icecream$puzzle, icecream$video)
```

You can also use the formula interface with `with()`:

```{r}
with(icecream, cor(puzzle, video))
```

### 1.2 Visualising the relationship

Base R plot version (quick and simple):

```{r}
plot(
  icecream$puzzle, icecream$video,
  main = "Correlation between puzzle score\nand video game score",
  xlab = "Puzzle score",
  ylab = "Video game score",
  sub  = paste0("r = ", round(cor(icecream$puzzle, icecream$video), 2))
)
```

> **Bonus task:**
> Recreate this plot using **ggplot2** instead of base R.

### 1.3 Non-parametric correlations

Sometimes variables are not normally distributed or contain outliers.
Two common **rank-based** alternatives are:

- **Spearman’s rho**
- **Kendall’s tau**

```{r}
# Spearman's rank correlation
cor(icecream$puzzle, icecream$video, method = "spearman")

# Kendall's tau
cor(icecream$puzzle, icecream$video, method = "kendall")
```

> **Reflection prompt:**  
> Compare the Pearson, Spearman, and Kendall coefficients.  
> Are they similar, or do ranks tell a different story?

---

## 2. Crime data: correlation matrices and tests

The `crime` dataset contains several numeric variables such as:

- `CrimeRate` – crime per population  
- `Males` – number of males in the labour force  
- `Wage` – average wage  
- `StateSize` – size of the state, and many others.

### 2.1 Simple correlation matrix

We can compute a correlation matrix for a small subset of variables:

```{r}
crime %>%
  select(CrimeRate, Males, Wage) %>%
  cor()
```

Rounded for nicer printing:

```{r}
crime %>%
  select(CrimeRate, Males, Wage) %>%
  cor() %>%
  round(2)
```

### 2.2 Pairwise scatterplots

A quick way to **visualise all pairwise relationships** is `pairs()`:

```{r}
crime %>%
  select(CrimeRate, Males, Wage) %>%
  pairs()
```

> **Bonus task:**
> Which pairs look most linear and which look weak or non-linear?

### 2.3 Testing correlations: `cor.test()`

`cor()` only gives you the coefficient.  
Use `cor.test()` when you want a **hypothesis test** and a **confidence interval**.

```{r}
# Test correlation between Males and Wage
cor.test(crime$Males, crime$Wage)

# Test correlation between Males and CrimeRate
cor.test(crime$Males, crime$CrimeRate)
```


---

## 3. Correlations by group

Sometimes you do not just want “one correlation for everybody”, but
separate correlations **within subgroups**.

In the `crime` data, variable `MoreMales` indicates whether the state
has a particularly high number of males.

### 3.1 Grouped correlation (summarise version)

Here we compute the correlation between `StateSize` and `Education`
**separately** for each value of `MoreMales`:

```{r}
crime %>%
  select(CrimeRate, StateSize, Education, Wage, MoreMales) %>%
  group_by(MoreMales) %>%
  summarise(
    cor_StateSize_Education = cor(StateSize, Education, use = "complete.obs")
  )
```

This approach is:

- readable,  
- uses only `dplyr`,  
- easy to adapt to other variable pairs.

### 3.2 Grouped correlation (nested + purrr version)

The same idea, but written in a more functional style:

```{r}

crime %>%
  select(CrimeRate, StateSize, Education, Wage, MoreMales) %>%
  group_by(MoreMales) %>%
  nest() %>%
  mutate(
    cor_StateSize_Education = map_dbl(
      data,
      ~ cor(.x$StateSize, .x$Education, use = "complete.obs")
    )
  )

```

- `nest()` creates a small tibble for each group in column `data`.  
- `map_dbl()` applies `cor()` inside each small dataset.

> This pattern is a gentle introduction to the “split–apply–combine” idea using **nested data frames**.

---

## 4. Visualising correlations in larger sets of variables

### 4.1 Correlation heatmap with `corrplot`

For many variables, a correlation matrix is easier to read as a
**heatmap**.

The next block:

1. Keeps only numeric variables.
2. Drops variables with missing values or zero variance.
3. Randomly picks 10 numeric variables (just for demo).
4. Computes a correlation matrix.
5. Plots it with `corrplot()` and hierarchical clustering.

```{r}
set.seed(123)  # for reproducible sampling of columns

crime %>%
  select_if(is.numeric) %>%               # keep only numeric variables
  select_if(~ !any(is.na(.))) %>%        # drop variables with any missing values
  select_if(~ sd(.) != 0) %>%            # drop variables with zero variance
  select(sample(names(.), size = 10)) %>%
  cor(use = "complete.obs") %>%          # row-wise removal of missing values
  corrplot::corrplot(order = "hclust")   # cluster similar variables
```

> **Bonus task:**
> Choose your own subset of variables and interpret at least three interesting pairs.

### 4.2 “Smart” correlation helpers from `lares`

The `lares` package provides convenience functions
for quickly exploring correlations in a dataset.

#### 4.2.1 Top correlated pairs

```{r}
crime %>%
  select(CrimeRate, Youth, Education, Males, StateSize, Wage, BelowWage) %>%
  lares::corr_cross(pvalue = 0.01, top = 10)
```

Same function, but **without plotting**, returning only a table:

```{r}
crime %>%
  select(CrimeRate, Youth, Education, Males, StateSize, Wage, BelowWage) %>%
  lares::corr_cross(pvalue = 0.01, top = 10, plot = FALSE)
```

#### 4.2.2 Correlations with a selected variable

You can also focus on correlations with one target variable,
for example `CrimeRate`:

```{r}
crime %>%
  select(CrimeRate, Youth, Education, Males, StateSize, Wage, BelowWage) %>%
  lares::corr_var(var = CrimeRate, top = 15)
```

