---
title: "Binomial distribution: coin flips"
code-fold: false
---

This page introduces the binomial distribution using coin flips:

* we start with simple rbinom() simulations
* then move to histograms and densities
* and finally compare simulated probabilities to the exact binomial values from
  dbinom() and pbinom().

The goal is to connect intuition (simulated coin flips) with formal
probability (binomial formulas).

## 1. Setup

```{r}
library(tidyverse)
library(flextable)

theme_colors <- data.frame(
    light_blue   = "#619CFF",
    light_orange = "#ffc461"
)

theme_colors

theme_set(theme_light())
```

## 2. Flipping a coin with rbinom()

rbinom() generates random outcomes from a binomial distribution:
• n = number of generated observations
• size = number of trials (e.g. number of flips)
• prob = probability of success on each trial

### 2.1 One coin, one flip

```{r}

# One Bernoulli trial: 0 = tails, 1 = heads

rbinom(n = 1, size = 1, prob = 0.5)
```

### 2.2 One coin, many flips

```{r}

# One coin flipped 10 times

rbinom(n = 1, size = 10, prob = 0.5)

# 10 separate flips of one coin (10 Bernoulli trials)

rbinom(n = 10, size = 1, prob = 0.5)
```

## 3. Many flips, many coins

Now we simulate a lot of flips to see how the distribution of heads behaves.

### 3.1 One coin flipped 100,000 times

```{r}

data.frame(
    flips_1 = rbinom(n = 100000, size = 1, prob = 0.5)
) %>%
    as_tibble() %>%
    rowid_to_column("id")

```

### 3.2 10, 100, and 1000 coins (each repeated 100,000×)

Each row below is one experiment (e.g. "flip 10 coins and count heads").
We repeat such experiments 100,000 times for 10, 100, and 1000 coins.

```{r}
coin_flips_multiple <- data.frame(
    flips_10   = rbinom(n = 100000, size = 10,   prob = 0.5),
    flips_100  = rbinom(n = 100000, size = 100,  prob = 0.5),
    flips_1000 = rbinom(n = 100000, size = 1000, prob = 0.5)
) %>%
    as_tibble() %>%
    rowid_to_column("id") %>%
    tidyr::pivot_longer(
        cols = starts_with("flips_"),
        names_to = "n",
        values_to = "heads"
    )

coin_flips_multiple
```

### 3.3 Empirical probabilities

```{r}
coin_flips_multiple_prob <- coin_flips_multiple %>%
    group_by(n, heads) %>%
    summarise(freq = n(), .groups = "drop") %>%
    mutate(prob = freq / 100000)

coin_flips_multiple_prob
```

## 4. Visualising the distributions

### 4.1 Histograms

```{r}
ggplot(coin_flips_multiple, aes(x = heads)) +
    geom_histogram(fill = "steelblue", bins = 30) +
    facet_wrap(~ n, scales = "free") +
    labs(
        x = "Number of heads",
        y = "Count",
        title = "Distribution of heads for 10, 100, and 1000 coin flips"
    )
```

### 4.2 Smoothed densities

```{r}
ggplot(coin_flips_multiple, aes(x = heads)) +
    geom_density(fill = "steelblue", alpha = 0.5, adjust = 10) +
    facet_wrap(~ n, scales = "free_x") +
    labs(
        x = "Number of heads",
        y = "Density",
        title = "Approximate density of heads counts"
    )
```

## 5. Exact distribution for 10 flips

Focus now on 10-coin experiments and summarise how often we see 0, 1, 2, …, 10
heads.

```{r}
coin_flips <- coin_flips_multiple %>%
    tidyr::pivot_wider(
        id_cols = id,
        names_from = n,
        values_from = heads
    )

coin_flips_results <- data.frame(
    table(coin_flips$flips_10),
    prop.table(table(coin_flips$flips_10)) %>% round(3),
    cumulative = cumsum(prop.table(table(coin_flips$flips_10))) %>% round(3)
) %>%
    select(-Var1.1)

names(coin_flips_results) <- c("result", "n", "prop", "cumulative")

coin_flips_results <- coin_flips_results %>%
    as.matrix() %>%
    t() %>%
    as.data.frame()

coin_flips_results <- rownames_to_column(coin_flips_results, "heads")

flextable::flextable(coin_flips_results[-1, ])
```

## 6. Mean and SD from simulation

Simulate 10-coin experiments and measure mean and standard deviation of heads.

```{r}
flips <- rbinom(n = 100000, size = 10, prob = 0.5)

sd(flips)
mean(flips)

# Sanity check: proportions must sum to 1

mean(flips == 5) + mean(flips < 5) + mean(flips > 5)
```

## 7. Visualising probabilities for 10 flips

### 7.1 Bar plot of empirical probabilities

```{r}
flips <- rbinom(n = 100000, size = 10, prob = 0.5)

flips %>%
    table() %>%
    as_tibble() %>%
    ggplot(aes(x = factor(., levels = 0:10), y = n / 100000, fill = .)) +
        geom_col() +
        theme(legend.position = "none", text = element_text(size = 20)) +
        xlab("Number of heads out of 10 coin flips") +
        ylab("Result probability") +
        ggtitle("Empirical distribution of heads (10 flips)")
```

### 7.2 Highlight exactly 5 heads

```{r}
flips %>%
    table() %>%
    as_tibble() %>%
    ggplot(aes(x = factor(., levels = 0:10), y = n / 100000, fill = !(. == 5))) +
        geom_col() +
        theme(legend.position = "none", text = element_text(size = 20)) +
        xlab("Number of heads out of 10 coin flips") +
        ylab("Result probability") +
        ggtitle("Highlighting exactly 5 heads out of 10")

mean(flips == 5)
```

### 7.3 Highlight "extreme" results (0–2 heads)

```{r}
flips %>%
    table() %>%
    as_tibble() %>%
    ggplot(aes(x = factor(., levels = 0:10), y = n / 100000, fill = !(factor(., levels = 0:10) %in% 0:2))) +
        geom_col() +
        theme(legend.position = "none", text = element_text(size = 20)) +
        xlab("Number of heads out of 10 coin flips") +
        ylab("Result probability") +
        ggtitle("Highlighting 0–2 heads out of 10")

mean(flips < 3)
sum(flips < 3)
```

## 8. Exact binomial probabilities

Now we move from simulation to exact formulas.

### 8.1 Point probabilities with dbinom()

```{r}
dbinom(x = 5,  size = 10, prob = 0.5)  # exactly 5 heads out of 10
dbinom(x = 6,  size = 10, prob = 0.5)  # exactly 6 heads
dbinom(x = 10, size = 10, prob = 0.5)  # all 10 heads
```

### 8.2 Cumulative probabilities with pbinom()

```{r}

# P(X <= 4) for X ~ Binomial(size = 10, prob = 0.5)

pbinom(q = 4, size = 10, prob = 0.5)
```

Compare with simulation:

```{r}
flips <- rbinom(n = 100000, size = 10, prob = 0.5)
mean(flips <= 4)

mean(rbinom(n = 100000, size = 10, prob = 0.5))
sd(rbinom(n = 100000, size = 10, prob = 0.5))
```

### 8.3 Theoretical mean and variance

For X \sim \text{Binomial}(n, p):

* \mathbb{E}(X) = np
* \text{Var}(X) = np(1 - p)

```{r}
size <- 100  # number of coins
prob <- 0.5  # probability of heads

data.frame(
    var  = round(size * prob * (1 - prob), 3),
    sd   = round(sqrt(size * prob * (1 - prob)), 3),
    mean = size * prob
)
```

## 9. Example: "Extreme" result with 100 coins

### 9.1 Example: How extreme is 56 heads out of 100?

We now switch from intuition (simulation) to exact probability using the binomial distribution
\(X \sim \mathrm{Binomial}(100, 0.5)\\).

The expected number of heads is:

* mean = 50
* standard deviation = $\sqrt{100 \cdot 0.5 \cdot 0.5} = 5$

So 56 heads is more than 1 SD above the mean, and values of 44 or fewer are symmetric on the left.

This motivates the idea of a two-tailed test:
“How surprising is it to be this far from the mean (on either side)?”

```{r}

dbinom(x = 56, size = 100, prob = 0.5)

```

The value `0.03895` means that **if the coin is truly fair**, the probability of getting
**exactly 56 heads out of 100 flips** is about **3.9%**.

In other words, if you repeated this experiment many times (always 100 flips of a fair coin),
you would expect an outcome of *exactly* 56 heads in roughly **4 out of 100 experiments**.
It’s rare, but not impossible — which is exactly the kind of “unusual but still plausible”
event that hypothesis tests are built around.

### 9.2 Two-tailed probability: 56 or more heads, OR 44 or fewer

We look at:

* right tail: 56, 57, …, 100
* left tail: 0, 1, …, 44
(because 50 – (56 – 50) = 44)

We now compute the probability of being at least this extreme in both tails:

```{r}
n <- 100
p <- 0.5

# P(X >= 56)

P_ge_56 <- sum(dbinom(x = 56:100, size = n, prob = p))

# P(X <= 44)

P_le_44 <- sum(dbinom(x = 0:44, size = n, prob = p))

# Two-tailed probability

P_ge_56 + P_le_44

```

::: {.callout-tip title="Interpreting the two-tailed probability"}

The value of **0.27** means that about **27 % of all outcomes** (under a perfectly fair coin) are **at least as extreme** as getting 56 heads out of 100 tosses.

### What does “as extreme” mean?

We consider outcomes that are **at least 6 away from the mean (50)**:

- **Right tail:** 56, 57, …, 100
- **Left tail:** 44, 43, …, 0
  (because 56 is 6 above the mean → we also check 6 below the mean)

This includes **many** results, not just a single number — which is why the probability is much larger than `P(X = 56)`.

### Why is 27 % not surprising?

Because in a two-tailed test we ask:

> **“How often would we see a result at least this far from the mean if the coin were fair?”**

The answer is:
👉 **Roughly one out of four experiments** will produce a result this extreme or more.

So **56 heads out of 100 is perfectly compatible with a fair coin**.
It is *not* evidence of bias.

:::


### 9.3 Visualising the extremeness

```{r}
# Compute the full binomial distribution
df <- data.frame(
  heads = 0:100,
  prob  = dbinom(0:100, size = 100, prob = 0.5)
)

df <- df %>%
  mutate(
    extreme = (heads >= 56) | (heads <= 44)   # highlight tail areas
  )

ggplot(df, aes(x = heads, y = prob, fill = extreme)) +
    geom_col() +
    scale_fill_manual(
        values = c("FALSE" = "grey80", "TRUE" = "#FC4E07"),
        labels = c("Typical outcomes", "Extreme outcomes")
    ) +
    labs(
        title = "Two-tailed extremeness for 100 coin flips",
        subtitle = "Values ≥ 56 or ≤ 44 are symmetric deviations from the mean (50)",
        x = "Number of heads",
        y = "Exact probability P(X = x)"
    ) +
        theme_minimal(base_size = 16) +
        theme(
        legend.title = element_blank(),
        legend.position = "bottom"
    )
```

### 9.4 Interpretation

::: {.callout-tip title="How to read this graph (interpretation)"}

This visualisation shows **exact probabilities** for all possible numbers of heads in 100 coin flips.

### What the shape tells us

- The **center** (around 50) holds most of the probability mass.
  This is what you expect from a fair coin: outcomes cluster tightly around the mean.

- The orange bars highlight the **extreme regions**:
  - **56, 57, …, 100** on the right
  - **44, 43, …, 0** on the left
  These values are equally “far from the mean” in either direction.

- Each orange bar individually is **rare**, and together they form the total probability of
  **“being at least this far from 50”**.

### What this means for our example

A result like **56 heads** is surprising — but **not shocking**.
About **27 %** of all experiments would produce an outcome at least this extreme
(either ≥ 56 or ≤ 44), *even if the coin is perfectly fair*.

In other words:

> **Unusual outcomes happen quite often.**
> A single result like 56/100 is not strong evidence that the coin is biased.

### Connection to hypothesis testing

This is exactly the logic behind **two-sided p-values**:
we evaluate how often we would observe a result **at least this extreme** under the assumption that nothing unusual is happening.

:::


