---
title: "Exploring the Apple Fitness Dataset"
code-fold: false
toc-depth: 3
---

## 1. About the dataset

This dataset contains **6,264 observations** collected from an Apple Watch and Fitbit-like device.  
Each row represents *one time slice* of recorded physical activity and biometric signals.

The variables describe:

- demographic attributes (`age`, `gender`, `height`, `weight`)
- physical activity metrics (`steps`, `distance`, `calories`)
- heart-related measurements (`hear_rate`, `resting_heart`, `entropy_heart`, etc.)
- computed features useful for modelling (`corr_heart_steps`, `steps_times_distance`)
- the recorded *activity class* (`activity`), e.g. `"Lying"`, `"Sitting"`, `"Walking"`, `"Running"`…

The dataset is ideal for demonstrating **linear regression**, **feature engineering**,  
and exploratory visualisation because its variables are intuitive and quantitatively meaningful.

---

## 2. Download the data

> **Download dataset (CSV):**
> [`apple_watch_fitbit_data.csv`](../assets/data/apple_watch_fitbit_data.csv)

---

## 3. Load packages and data

```{r}

library(broom)
library(modelr)
library(tidymodels)
library(tidyverse)

apple <- read_csv("../assets/data/apple_watch_fitbit_data.csv")

glimpse(apple)
```
## 4. Basic structure of the dataset

The output of glimpse() shows 18 variables:

* age, gender, height, weight
* steps, hear_rate, distance, calories
* entropy_heart, entropy_setps
* resting_heart, norm_heart, sd_norm_heart
* intensity_karvonen
* corr_heart_steps (correlation between steps & heart rate over a time window)
* steps_times_distance (engineered feature)
* device (factor)
* activity (categorical label)

We will use these variables to introduce linear regression step by step — from simple models to more advanced ones (interactions, multivariate regression, transformations, diagnostics, etc.).

## 5. First quick check: distributions

```{r}
apple %>%
  select(steps, calories, hear_rate, distance) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "#619CFF", color = "white") +
  facet_wrap(~name, scales = "free") +
  theme_minimal(base_size = 14)
```

## Next steps

In the following sections we will:

1. Build the **first simple linear model** (calories ~ steps)
2. Explore **meanings of coefficients**
3. Add **multiple predictors**
4. Diagnose the model (residuals, QQ plot)
5. Try an **interaction term** and see how interpretation changes
6. Reflect on *when linear regression stops making sense

## Packages for regression in this chapter

In this chapter we will work with a small, focused toolbox for regression models in R.

- **Base `stats` (`lm()` function)**
  Simple linear and multiple regression using base R. We use it as the core engine:
  `lm(hear_rate ~ steps, data = apple)` is our starting point for modelling.

- **`broom`**
  Tidies model output into tibbles.
  - `tidy()` – coefficients in a neat table,
  - `augment()` – fitted values and residuals attached to the original data,
  - `glance()` – one-row summary (R², AIC, …).
  This makes regression results easy to combine with `dplyr` and `ggplot2`.

- **Optional: `modelr`**
  Helper functions like `add_predictions()` and `add_residuals()`.
  We will mostly rely on `broom`, but you may encounter `modelr` in other tutorials.

- **Optional: `tidymodels`**
  A modern modelling framework with a unified grammar for regression, classification,
  resampling, and tuning.
  We briefly show how `linear_reg()` with engine `"lm"` reproduces the basic `lm()` workflow,
  and how the same interface extends to more complex models.

- **Optional: diagnostics packages (`performance`, `car`)**
  For this R-focused course we keep diagnostics simple (plots and a few summary numbers).
  For deeper regression diagnostics (multicollinearity, detailed ANOVA, assumption checks),
  packages like `performance` or `car` are good next steps, especially in a statistics-oriented course.



## 6. First simple model: calories ~ steps

A natural first question:

How strongly are burned calories related to the number of steps?

We start with a simple linear regression:

```{r}
model_simple <- lm(calories ~ steps, data = apple)
model_simple
```

The printed output shows the intercept and slope, but is not very tidy.
We can use broom::tidy() instead:

```{r}
tidy(model_simple)
```

* the intercept is the expected calories when steps = 0,
* the slope tells us how many calories increase on average when we add one step.

(Because steps can be large, it may be more interpretable to rescale them, for example to
thousands of steps, later in the chapter.)

### 6.1 Visualising the simple regression line

```{r}

apple %>%
  ggplot(aes(x = steps, y = calories)) +
  geom_point(alpha = 0.3, color = "#619CFF") +
  geom_smooth(method = "lm", se = TRUE, color = "#F8766D") +
  labs(
    title = "Simple linear regression: calories ~ steps",
    x = "Steps",
    y = "Calories"
  ) +
  theme_minimal(base_size = 14)
```

The regression line shows the average trend; the scatter of points around the line
represents unexplained variability.

⸻

## 7. Using broom for a tidy workflow

### 7.1 Coefficient table

```{r}
coef_table <- tidy(model_simple)
coef_table
```

This gives a tibble with:
* estimate – coefficient values,
* std.error, statistic, p.value – classical inferential quantities.

### 7.2 Model-level summary

```{r}
glance(model_simple)
```

Here we see:
* r.squared – proportion of variance in calories explained by steps,
* adj.r.squared – adjusted version,
* sigma – residual standard deviation,
* and other fit statistics.

### 7.3 Augmenting the data with fitted values and residuals

```{r}
apple_aug <- augment(model_simple, data = apple)
glimpse(apple_aug)
```


New columns:
* .fitted – predicted calories from the model,
* .resid – residuals (observed − fitted).

We can use them for diagnostic plots.

⸻

## 8. Basic diagnostics for the simple model

### 8.1 Residuals vs fitted values

```{r}
apple_aug %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(alpha = 0.3, color = "#619CFF") +
  labs(
    title = "Residuals vs fitted values",
    x = "Fitted calories",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 14)
```

We would like to see:
* roughly constant spread of residuals across the range of fitted values,
* no strong curvature.

### 8.2 Histogram and Q–Q plot of residuals

```{r}
# Histogram
apple_aug %>%
  ggplot(aes(x = .resid)) +
  geom_histogram(bins = 40, fill = "#619CFF", color = "white") +
  labs(
    title = "Distribution of residuals",
    x = "Residual",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

# Q-Q plot
apple_aug %>%
  ggplot(aes(sample = .resid)) +
  stat_qq(color = "#619CFF") +
  stat_qq_line(color = "#F8766D") +
  labs(
    title = "Q-Q plot of residuals",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  theme_minimal(base_size = 14)
```

These plots help us visually assess the normality assumption of residuals.

::: callout-note

Interpretation reminder

For this course we keep diagnostics informal and visual.
In a full statistics course, you would add formal tests (normality, homoscedasticity) and
check for influential points.
:::

⸻

## 9. Multiple regression: adding distance and heart rate

Steps are not the only driver of calories. We can add more predictors:

```{r}
model_multi <- lm(calories ~ steps + distance + hear_rate, data = apple)

tidy(model_multi)
```



The interpretation now becomes:
* each coefficient is a partial effect,
* for example, the slope for steps describes the effect of steps when distance and heart rate are held constant.

Compare overall fit:

glance(model_simple)[c("r.squared", "adj.r.squared")]
glance(model_multi)[c("r.squared", "adj.r.squared")]

::: callout-question

Check your understanding
* Did the multiple regression improve $R^2$ markedly?
* Are all predictors statistically significant?
* Would you keep all of them in a parsimonious model?

Try to answer these questions before scrolling further or printing the summary in R.
:::

⸻


## 10. Visualising model predictions

To keep the plot in two dimensions, we can fix some variables and vary only one of them.
For example, fix hear_rate at its median and see how predicted calories grow with steps.

```{r}

hear_median <- median(apple$hear_rate, na.rm = TRUE)

pred_grid <- tibble(
  steps     = seq(min(apple$steps), max(apple$steps), length.out = 100),
  distance  = median(apple$distance, na.rm = TRUE),
  hear_rate = hear_median
)

pred_grid <- pred_grid %>%
  mutate(calories_pred = predict(model_multi, newdata = pred_grid))

ggplot() +
  geom_point(
    data = apple,
    aes(x = steps, y = calories),
    alpha = 0.2,
    color = "#B3CDE3"
  ) +
  geom_line(
    data = pred_grid,
    aes(x = steps, y = calories_pred),
    color = "#F8766D",
    linewidth = 1.2
  ) +
  labs(
    title = "Predicted calories as a function of steps",
    subtitle = "Distance and heart rate fixed at their medians",
    x = "Steps",
    y = "Calories"
  ) +
  theme_minimal(base_size = 14)
```

This connects the algebraic model to a visual, more intuitive understanding.

⸻

## 11. Optional: interaction term

You might suspect that the effect of steps is different for low vs high heart rate.
This can be expressed by an interaction:

model_interaction <- lm(calories ~ steps * hear_rate + distance, data = apple)
tidy(model_interaction)

The term steps:hear_rate captures how the slope of steps changes with heart rate.

::: callout-tip

When to use interactions?

Use interaction terms when:
* you expect the effect of one predictor to depend on the level of another predictor,
* and you can explain such dependence substantively (not only because a p-value is small).
:::

⸻

## 11. Optional: the same model via tidymodels

This course focuses on lm(), but it is useful to see that the same idea fits into the
tidymodels ecosystem.

```{r}

lm_spec <- linear_reg() %>%
  set_engine("lm")

lm_fit <- lm_spec %>%
  fit(calories ~ steps + distance + hear_rate, data = apple)

lm_fit

# We can extract a tidy summary using:
tidy(lm_fit)
```

Later, the same interface can be reused for regularised regression,
tree-based models, and more — without changing the high-level syntax.

⸻

## 12. Summary

In this chapter you have:

* loaded and explored the Apple fitness dataset,
* fitted a simple regression calories ~ steps,
* interpreted slope, intercept, and $R^2$,
* used broom to obtain tidy model summaries,
* inspected residuals visually,
* extended the model to multiple predictors and an optional interaction,
* and seen how the same model looks in tidymodels.

These are the core ingredients needed for building and understanding
linear regression models in R. More advanced topics (regularisation, cross-validation,
non-linear models) can be layered on top of this foundation.