---
title: "Spam detection (Supervised)"
subtitle: "Logistic regression + decision trees: classification & interpretation"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

## Goal of this section

Here we treat spam detection as a **supervised classification** problem:

- we have features extracted from emails (predictors)
- we have a label (spam / not spam)
- we train a model on a training set
- we evaluate on a test set (data the model did not see)

We focus on **interpretation** as much as on prediction:

- What does a coefficient mean in logistic regression?
- How do decision trees make decisions?
- What does a confusion matrix actually tell us?
- Why accuracy alone can be misleading?

## Dataset

- Dataset: `spam.txt`  
[Download dataset](../assets/data/spam_detection_data.txt)

Assumption used below:
- last column is the label (TRUE / FALSE)
- all other columns are numeric predictors

## Setup

```{r}
library(tidyverse)

# modelling
library(rpart)
library(rpart.plot)

# evaluation
library(yardstick)
library(broom)
```

## Load data

```{r}
spam_raw <- read_csv2("../assets/data/spam_detection_data.csv", col_names = TRUE)

spam <- spam_raw %>%
  mutate(
    row_id = row_number(),
    label  = as.factor(yesno)
  ) %>%
  select(-yesno)

# sanity checks (nepřeskakuj!)
stopifnot("label" %in% names(spam))
nlevels(spam$label)
table(spam$label)

glimpse(spam_raw)
summary(spam_raw[, 1:6])
```

Create a clean data frame with a factor label.

```{r}
spam <- spam_raw %>%
  rename(label = !!sym(names(spam_raw)[ncol(spam_raw)])) %>%
  mutate(label = factor(label)) %>%
  # ensure numeric predictors
  mutate(across(-label, as.numeric))

table(spam$label)
```

## Train / test split

We use a **stratified split** to make sure both classes (spam / not spam) are present in the training set.
This prevents rare edge cases where a random split accidentally puts almost all positives into test (or train),
which can break some models (notably `rpart` probability predictions).

```{r}
set.seed(1)
train_frac <- 0.7

idx_train <- spam %>%
  mutate(i = row_number()) %>%
  group_by(label) %>%
  slice_sample(prop = train_frac) %>%
  ungroup() %>%
  pull(i)

train <- spam[idx_train, ]
test  <- spam[-idx_train, ]

nrow(train); nrow(test)
table(train$label)
table(test$label)
```


## Baseline (always predict the majority class)

```{r}
majority_class <- names(sort(table(train$label), decreasing = TRUE))[1]
majority_class
```

```{r}
majority_class <- names(sort(table(train$label), decreasing = TRUE))[1]

baseline_pred <- factor(rep(majority_class, nrow(test)), levels = levels(train$label))

yardstick::accuracy_vec(truth = test$label, estimate = baseline_pred)

baseline_tbl <- tibble(truth = test$label, estimate = baseline_pred)
yardstick::conf_mat(baseline_tbl, truth, estimate)
```

::: callout-tip
### Why we start with a baseline
If your “fancy model” cannot beat the baseline, something is wrong:
- data leakage, coding mistake, or
- the features contain little signal.
:::

## Model 1: Logistic regression

### Fit

We use a standard logistic regression (`glm(..., family = binomial)`).
This model estimates:

- **log-odds** of the positive class as a linear function of predictors
- which can be converted to predicted probabilities

```{r}
fit_glm <- glm(label ~ ., data = train, family = binomial)
summary(fit_glm)
```

### Predicted probabilities and class prediction

```{r}
p_glm <- predict(fit_glm, newdata = test, type = "response")
head(p_glm)
```

Choose a decision threshold (default: 0.5).

```{r}
threshold <- 0.5
pred_glm <- ifelse(p_glm >= threshold, levels(train$label)[2], levels(train$label)[1])
pred_glm <- factor(pred_glm, levels = levels(train$label))
```

### Evaluation

```{r}
glm_tbl <- tibble(truth = test$label, estimate = pred_glm)
yardstick::conf_mat(glm_tbl, truth, estimate)
yardstick::accuracy_vec(truth = test$label, estimate = pred_glm)
yardstick::sens_vec(truth = test$label, estimate = pred_glm)  # sensitivity / recall
yardstick::spec_vec(truth = test$label, estimate = pred_glm)  # specificity
yardstick::f_meas_vec(truth = test$label, estimate = pred_glm) # F1 score
```

### Interpretation: coefficients

A coefficient tells you how the predictor changes the **log-odds** of the positive class
while holding other predictors constant.

To make this more intuitive, we can also look at **odds ratios**.

```{r}
coef_table <- broom::tidy(fit_glm) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  arrange(desc(abs(estimate)))

coef_table %>%
  select(term, estimate, std.error, statistic, p.value, odds_ratio) %>%
  print(n = 15)
```

::: callout-note
### Practical reading tip
- `odds_ratio > 1` increases odds of the positive class
- `odds_ratio < 1` decreases odds  
But: odds ratios are *not* probability changes, and correlated predictors can distort interpretation.
:::

## Model 2: Decision tree

### Fit

```{r}
fit_tree <- rpart(label ~ ., data = train, method = "class")
fit_tree
```

### Plot

```{r}
rpart.plot(fit_tree, type = 3, extra = 104, fallen.leaves = TRUE)
```

Interpretation reminder:
- follow a path from the root to a leaf
- each split is a rule
- leaves show class proportions
- always check support (how many cases reach a leaf)

### Predict and evaluate

In some `rpart` versions/data combinations, `predict(..., type = "prob")` can throw an internal error.
For teaching purposes, **`type = "class"` is the most robust** and is enough for confusion matrices and metrics.

```{r}
pred_tree <- predict(fit_tree, newdata = test, type = "class")
pred_tree <- factor(pred_tree, levels = levels(train$label))

tree_tbl <- tibble(truth = test$label, estimate = pred_tree)
yardstick::conf_mat(tree_tbl, truth, estimate)
yardstick::accuracy_vec(truth = test$label, estimate = pred_tree)
yardstick::f_meas_vec(truth = test$label, estimate = pred_tree)
```

(Optional) If you really want probabilities, try this chunk. If it fails on your system, just skip it.

```{r}
p_tree <- tryCatch(
  predict(fit_tree, newdata = test, type = "prob"),
  error = function(e) NULL
)

if (!is.null(p_tree)) {
  head(p_tree)
}
```

## Comparing models

```{r}
metrics <- tibble(
  model = c("Baseline", "Logistic regression", "Decision tree"),
  accuracy = c(
    yardstick::accuracy_vec(test$label, baseline_pred),
    yardstick::accuracy_vec(test$label, pred_glm),
    yardstick::accuracy_vec(test$label, pred_tree)
  ),
  f1 = c(
    yardstick::f_meas_vec(test$label, baseline_pred),
    yardstick::f_meas_vec(test$label, pred_glm),
    yardstick::f_meas_vec(test$label, pred_tree)
  )
)

metrics
```

::: callout-warning
### Accuracy is not enough
If the classes are imbalanced, a model can have high accuracy while being useless.
Always inspect:
- confusion matrix
- sensitivity/recall (how many positives you catch)
- specificity (how many negatives you correctly reject)
- F1 score (balance of precision/recall)
:::

## Threshold tuning (optional but useful)

The 0.5 threshold is arbitrary.
Depending on the cost of mistakes, you might prefer to catch more spam (higher sensitivity)
even at the cost of more false positives.

```{r}
threshold_grid <- seq(0.1, 0.9, by = 0.05)

threshold_results <- map_dfr(threshold_grid, function(t) {
  pred <- factor(ifelse(p_glm >= t, levels(train$label)[2], levels(train$label)[1]),
                 levels = levels(train$label))
  tibble(
    threshold = t,
    accuracy = yardstick::accuracy_vec(test$label, pred),
    sens = yardstick::sens_vec(test$label, pred),
    spec = yardstick::spec_vec(test$label, pred),
    f1 = yardstick::f_meas_vec(test$label, pred)
  )
})

threshold_results %>%
  pivot_longer(cols = c(accuracy, sens, spec, f1),
               names_to = "metric", values_to = "value") %>%
  ggplot(aes(threshold, value, color = metric)) +
  geom_line() +
  geom_point() +
  labs(title = "Logistic regression: metrics vs threshold") +
  theme_light()
```

## Exercises

::: {.callout-tip collapse="true"}
### Exercise 1: Baseline vs model

- Compute the baseline accuracy (already done above).
- Which model beats it? By how much?
- Write one sentence explaining why baseline matters.
:::

::: {.callout-tip collapse="true"}
### Exercise 2: Interpret one logistic coefficient

Pick one predictor with a large absolute coefficient.
- What is its sign?
- What does that imply about odds?
- Compute and interpret its odds ratio.
:::

::: {.callout-tip collapse="true"}
### Exercise 3: Read a decision-tree rule

Pick one leaf in the tree and write the rule as:

> IF ... AND ... AND ... THEN predicted class = ... (with probability ≈ ...)

Also report how much data reaches that leaf (support).
:::

::: {.callout-tip collapse="true"}
### Exercise 4: Change the threshold

Try thresholds 0.3 and 0.7 for the logistic regression.
- Compare confusion matrices.
- Which threshold would you prefer if missing spam is costly?
- Which if false alarms are costly?
:::

## Solutions (brief)

::: {.callout-warning collapse="true"}
### Exercise 2 (solution sketch)

A positive coefficient increases **log-odds** of the positive class.
`exp(coef)` gives the odds ratio: values above 1 increase odds, below 1 decrease odds.
Interpretation is conditional on other predictors and can be distorted by collinearity.
:::

::: {.callout-warning collapse="true"}
### Exercise 3 (solution sketch)

A path from root to leaf is a conjunction of split conditions.
The leaf prediction is the majority class in that region; the probability shown is the class proportion within that leaf.
Small support means the rule is “sharp” but applies to few cases.
:::
