---
title: "Spam detection (Unsupervised)"
subtitle: "PCA + clustering: when we ignore the label"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

## Goal of this section

In the supervised part, we trained classifiers using known labels.
Here we do the opposite: we **pretend that the label does not exist** and try to
discover structure in the data using unsupervised methods.

The goal is not performance, but understanding:

- why **scaling** is crucial for unsupervised methods,
- how **PCA** helps us *see* high-dimensional data,
- why clustering results are always **hypotheses**, not ground truth.

## Dataset

We work with a classic spam dataset where:

- rows correspond to individual emails
- columns are numeric text features (e.g., frequencies of certain tokens)
- the last column is a binary label (spam / not spam)

In this section, the label will be used **only for evaluation after the fact**.

### Download

- Dataset: `spam_detection_data.txt`

[Download dataset](../assets/data/spam_detection_data.txt)

## Setup

```{r}
library(tidyverse)
library(cluster)    # silhouette
library(factoextra) # PCA / clustering helpers
```

## Loading the data

```{r}
spam_raw <- read_csv2("../assets/data/spam_detection_data.csv", col_names = TRUE)

spam_raw <- spam_raw %>% mutate(row_id = row_number())

glimpse(spam_raw)
summary(spam_raw[, 1:6])
```

Extract the label (last column) and inspect its distribution.

```{r}

label_df <- spam_raw %>%
  transmute(row_id, label = dplyr::pull(., ncol(.) - 1)) %>%  # last column before row_id
  mutate(label = factor(label))

table(label_df$label)
```

## Creating an unsupervised dataset

We now remove the label and keep only numeric features.  
Important: after cleaning (dropping rows / columns), we must keep the label **aligned** with the remaining rows.

```{r}
# 2) Label (in your data it is 'yesno' TRUE/FALSE)
label_df <- spam_raw %>%
  transmute(row_id, label = factor(yesno))

# 3) Predictors only (drop label, keep row_id)
X_df <- spam_raw %>%
  select(-yesno)    # keep row_id!

# 4) Make numeric matrix for PCA/k-means (drop row_id)
X_mat <- X_df %>%
  select(-row_id) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# 5) Remove rows with any non-finite values (NA/NaN/Inf)
keep_rows <- apply(is.finite(X_mat), 1, all)
X_mat <- X_mat[keep_rows, , drop = FALSE]
row_id_kept <- X_df$row_id[keep_rows]

# 6) Remove zero-variance columns (otherwise scaling creates NA)
sds <- apply(X_mat, 2, sd)
keep_cols <- is.finite(sds) & sds > 0
X_mat <- X_mat[, keep_cols, drop = FALSE]

# 7) Scale
X_scaled <- scale(X_mat)
stopifnot(all(is.finite(X_scaled)))

# 8) Align label to remaining rows
label_kept <- label_df$label[match(row_id_kept, label_df$row_id)]
stopifnot(length(label_kept) == nrow(X_scaled))
```
## PCA: visualizing structure

```{r}
pca <- prcomp(X_scaled, center = TRUE, scale. = FALSE)
summary(pca)
```

Projection onto the first two principal components:

```{r}
pca_df <- as_tibble(pca$x[, 1:2]) %>%
  mutate(label = label_kept)

ggplot(pca_df, aes(PC1, PC2, color = label)) +
  geom_jitter(alpha = 0.7, size = 2) +
  labs(
    title = "PCA projection (PC1 vs PC2)",
    subtitle = "True labels shown only for post-hoc inspection"
  ) +
  theme_light()
```

Interpretation:
PCA does not find clusters. It finds directions of maximal variance.
If spam and non-spam separate here, the signal is strong.
If not, the difference may be weak or hidden in higher dimensions.

## k-means clustering

### k = 2 (spam vs not spam hypothesis)

```{r}
set.seed(1)
km2 <- kmeans(X_scaled, centers = 2, nstart = 25)

table(cluster = km2$cluster, label = label_kept)
```

### Silhouette analysis

```{r}
set.seed(1)
idx <- sample(seq_len(nrow(X_scaled)), 300)

sil_small <- silhouette(km2$cluster[idx], dist(X_scaled[idx, ]))
plot(sil_small)

```

### Choosing k systematically

```{r}
k_grid <- 2:8

sil_scores <- map_dbl(k_grid, function(k) {
  km <- kmeans(X_scaled, centers = k, nstart = 25)
  mean(silhouette(km$cluster, dist(X_scaled))[, 3])
})

tibble(k = k_grid, silhouette = sil_scores) %>%
  ggplot(aes(k, silhouette)) +
  geom_line(size = 2) +
  geom_point(size = 3) +
  theme_light() +
  labs(title = "Silhouette score vs number of clusters")
```

## Exercises

::: {.callout-tip collapse="true"}
### Exercise 1: What happens without scaling?

- Run PCA on `X` instead of `X_scaled`.
- Compare explained variance and the PC1–PC2 plot.
- Explain the difference in two sentences.
:::

::: {.callout-tip collapse="true"}
### Exercise 2: Are clusters the same as labels?

- For k = 2, inspect `table(cluster, label)`.
- Compute a simple cluster purity.
- Explain why this is not the same as classification accuracy.
:::

::: {.callout-tip collapse="true"}
### Exercise 3: Choosing k

- Find the k that maximizes the silhouette score.
- Inspect the cluster–label table for this k.
- Discuss whether a higher silhouette always means a better solution
  for the spam detection problem.
:::

## Solutions (brief)

::: {.callout-warning collapse="true"}
### Exercise 1 (solution sketch)

Without scaling, variables with larger variance dominate both PCA and k-means.
Standardization ensures that all features contribute comparably, regardless of their original scale.
:::
