---
title: "Labelled scatterplot (Gapminder)"
---

This task focuses on adding readable labels to a scatterplot.  
You will use `geom_text_repel()` from the **ggrepel** package to avoid overlaps.

## Goal
Replicate a labelled scatterplot of **GDP per capita vs. life expectancy** for European countries in 2007.  
Focus first on *structure* (layers), then on *label readability*, and only at the end on *style*.

## Data
We use a prepared Gapminder subset saved as `src/data/gapminder.csv`.  
If you want to reproduce the file, see `ggplot-gallery/plot_prep/gapminder_labeled_scatter.R`.

<p>
  <a href="../assets/data/gapminder.csv" download>
    Download the dataset (gapminder.csv)
  </a>
</p>

```{r setup}
library(tidyverse)
library(ggrepel)
library(gapminder)
```

```{r data}
# Load prepared dataset
gapminder::gapminder

# Keep a readable subset for labelling
df <- gapminder %>%
  filter(year == 2007, continent == "Europe") %>%
  drop_na()
```

## Step 1: Basic scatterplot

```{r step1}
ggplot(df, aes(gdpPercap, lifeExp)) +
  geom_point()
```

What to notice:
- `aes(x, y)` maps variables to axes.
- `geom_point()` draws the observations.

## Step 2: Make points easier to read

```{r step2}
ggplot(df, aes(gdpPercap, lifeExp)) +
  geom_point(size = 2, alpha = 0.8)
```

Hints:
- `size` increases visibility.
- `alpha` adds transparency (helps if points overlap).

## Step 3: Add country labels (ggrepel)

```{r step3}
ggplot(df, aes(gdpPercap, lifeExp, label = country)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_text_repel(size = 3)
```

Hints:
- `label = country` tells ggplot what text to show.
- `geom_text_repel()` spreads labels so they do not collide.

## Step 4: Improve the x-axis scale

```{r step4}
ggplot(df, aes(gdpPercap, lifeExp, label = country)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_text_repel(size = 3) +
  scale_x_log10()
```

Notes:
- GDP per capita is right‑skewed.
- A log scale makes distances more meaningful.

## Step 5: Final polish (labels + clean theme)

Hints:
- add informative axis labels and title  
- keep all text at 14 pt  
- use a minimal theme for readability

```{r step5}
ggplot(df, aes(gdpPercap, lifeExp, label = country)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_text_repel(size = 3, max.overlaps = Inf) +
  scale_x_log10() +
  labs(
    x = "GDP per capita (log scale)",
    y = "Life expectancy (years)",
    title = "Gapminder 2007: Europe",
    subtitle = "Each point is one country"
  ) +
  theme_minimal(base_size = 14)
```

---

## Hidden solution

```{r solution}
ggplot(df, aes(gdpPercap, lifeExp, label = country)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_text_repel(size = 3, max.overlaps = Inf) +
  scale_x_log10() +
  labs(
    x = "GDP per capita (log scale)",
    y = "Life expectancy (years)",
    title = "Gapminder 2007: Europe",
    subtitle = "Each point is one country"
  ) +
  theme_minimal(base_size = 14)
```

## Extra challenges
1. Label only the **5 richest** and **5 poorest** countries.
2. Color points by life expectancy.
3. Try a different continent or a different year.
