---
title: "Gapminder Bubble Animation with Plotly"
code-fold: false
---

## Step 1 — Load packages and data

```{r}
library(gapminder)
library(plotly)
library(dplyr)

?gapminder

head(gapminder)
```

## Step 2 — Build animated bubble chart

```{r}

plot_ly(
  gapminder,
  x = ~gdpPercap,
  y = ~lifeExp,
  size = ~pop,
  color = ~continent,
  frame = ~year,
  text = ~paste(
    "Country:", country,
    "<br>Life expectancy:", lifeExp,
    "<br>GDP per capita:", round(gdpPercap),
    "<br>Population:", format(pop, scientific = FALSE)
  ),
  hoverinfo = "text",
  sizes = c(5, 80),
  type = "scatter",
  mode = "markers",
  colors = c("#EE4244FF", "#F8D961FF", "#B6D944FF", "#638E6EFF", "#3C5541FF", "#132157FF"),
  marker = list(opacity = 0.7, sizemode = "diameter")
) %>%
  layout(
    title = "Global Development Over Time",
    xaxis = list(title = "GDP per capita", type = "log"),
    yaxis = list(title = "Life expectancy"),
    legend = list(title = list(text = "<b>Continent</b>"))
  )
```

## Step 3 — Notes

* Each frame corresponds to one year in the dataset.
* Bubble size encodes population.
* Bubble color shows continent.
* The x-axis is on a log scale because GDP per capita varies widely.
* You can explore how health and wealth evolved globally from 1952–2007.


::: callout-question
## Did you notice anything interesting in 1992?

Have a look at the animation — does anything look unusual?

::: callout-body
::: {.callout-tip collapse="true"}
## Explanation

The dramatic collapse around **1992–1994** in Rwanda, Africa is real data — not an artefact.

It reflects the **Rwandan genocide (1994)** and the preceding civil war.
In the span of a few months, roughly **800,000 people** were killed, the health system collapsed, and mortality skyrocketed.

Gapminder uses 5-year intervals (1987 → 1992 → 1997), so the destruction of 1994 appears as a sharp fall in the **1992** data point.

This is one of the clearest examples where historical events leave a visible fingerprint in demographic data.

You can also tell that **Rwanda is not the only country with a story** hidden in the dots:

- Several **southern African countries** (e.g., Botswana, Zimbabwe, South Africa) show a fall in life expectancy in the 1990s associated with the **HIV/AIDS epidemic**.
- Some **post-Soviet countries** show stagnation or even decreases in life expectancy and income after the **collapse of the USSR**.
- Other countries (for example in East Asia) show long, smooth climbs in income and life expectancy that correspond to periods of **rapid economic development**.

:::
:::
:::

::: callout-question
## Are GDP per capita and life expectancy related?


::: callout-body
::: {.callout-tip collapse="true"}
## Explanation

The overall Pearson correlation in the full Gapminder dataset is
**r ≈ 0.58 (p < .001)**, with a confidence interval roughly **0.55–0.61**.

```{r}
cor.test(gapminder$lifeExp, gapminder$gdpPercap)

ggplot(gapminder, aes(x = lifeExp, y = gdpPercap)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~continent) +
    theme_light()
```

This means:

- Countries with **higher income per person** tend to have **longer life expectancy**.
- The relationship is strong, but not perfect – several regions rise faster or slower depending on history, politics, healthcare, or conflict.
- In the animation, this appears as a **general upward-right trend**: bubbles move toward higher GDP and higher life expectancy over time.

Interpreting this correlation requires some care:

- GDP per capita does *not* biologically “cause” longer life.
- Instead, it serves as a **proxy** for infrastructure, stability, healthcare availability, sanitation, education, and reduced child mortality, all of which increase average lifespan.
- Large deviations (e.g., sudden drops in Rwanda or HIV-related declines in southern Africa) reveal how **historical shocks** can break this pattern.

So while r ≈ 0.58 might look like a simple number, in context it summarises decades of global public-health, economics, and political history.
:::


::: callout-question
## Bonus task – per-continent correlation

Can you compute the correlation between GDP per capita and life expectancy **separately for each continent**?

Hint: Instead of writing five separate `cor.test()` calls, try to use a *grouped* workflow and let R repeat the test for you.

::: callout-body
::: {.callout-tip collapse="true"}
## Suggested solution

One neat approach is to combine:

- `group_by(continent)` – to split the data into continent-wise groups
- `nest()` – to store each group’s rows in a list-column `data`
- `map()` – to run `cor.test()` on each nested dataset
- `broom::tidy()` – to turn the test output into a clean tibble

```{r}
library(broom)
library(tidyverse)

gapminder %>%
  group_by(continent) %>%
  nest() %>%
  mutate(
    cor_test = map(
      data,
      ~ cor.test(.x$lifeExp, .x$gdpPercap, use = "complete.obs")
    ),
    tidy_cor = map(cor_test, broom::tidy)
  ) %>%
  select(continent, tidy_cor) %>%
  unnest(tidy_cor)
```
This way you get a tidy table with one row per continent, including the
correlation estimate, confidence interval, and p-value.


:::
:::
:::

