---
title: "3D regression surface with plotly"
code-fold: false
---

This page demonstrates how to build a **3D regression surface** using:

- two predictors: `Sepal.Width` and `Sepal.Length`
- one continuous outcome: `Petal.Width`
- a non-linear **support vector regression (SVR)** model
- an interactive **plotly** surface to explore the fitted function

The goal is to show how a smooth non-linear model passes through the observed
flower measurements in the classic `iris` dataset.

## Packages used in this example

**kernlab**
Provides kernel-based machine learning methods, including Support Vector
Machines.
Used here to fit a radial-basis-function SVR model.
<https://cran.r-project.org/package=kernlab>

**pracma**
Numerical tools for working with matrices and grids.
`meshgrid()` from this package generates the 2D prediction grid.
<https://cran.r-project.org/package=pracma>

**plotly**
Renders interactive 3D graphics—rotation, zooming, hover labels, and smooth
surface interpolation.
Core engine for the resulting visualisation.
<https://plotly.com/r/>

**tidymodels**
Unified modelling interface. Wraps the SVR model and simplifies fitting and
predicting.
<https://www.tidymodels.org/>

**tidyverse**
Provides clean data manipulation and piping. Used to select variables, create
grids, and process predictions before plotting.

The examples on this page use the classic `iris` dataset from the
**datasets** package in base R.

Each row describes one iris flower, measured by:

- **Sepal length / width** – size of the outer leaf-like part of the flower
- **Petal length / width** – size of the inner coloured petals
- **Species** – Setosa, Versicolor, or Virginica

The schematic below shows where *sepal* and *petal* are on the flower.

## Load data and packages

```{r}
library(kernlab)
library(pracma)
library(plotly)
library(tidymodels)
library(tidyverse)

(iris <- as_tibble(iris))

```

## Step 1 — Select predictors and outcome

We begin by choosing which variables define the **horizontal plane** of our 3D model and which variable we want to **predict**.

- `X` are the **predictors**.
  These two variables define the *input space* — every point in this plane will receive a predicted value.

- `y` is the **outcome**.
  This is the variable whose values we try to approximate with a smooth model.

Together they specify the regression problem:

> Predict *Petal.Width* from *Sepal.Width* and *Sepal.Length*.

```{r}
X <- iris %>%
  select(Sepal.Width, Sepal.Length)

y <- iris %>%
  select(Petal.Width)

X
y
```

```{r}
X <- iris %>%
  select(Sepal.Width, Sepal.Length)

X

y <- iris %>%
  select(Petal.Width)

y

```

## Step 2 — Understanding the SVR model

In this step we train a **Support Vector Regression (SVR)** model to learn the relationship:

$$
\text{Petal.Width} = f(\text{Sepal.Width},\ \text{Sepal.Length})
$$

The goal is not classification, but a **smooth, flexible regression surface** that later becomes the 3D plot.

---

### Support Vector Regression (SVR) in simple terms

SVR is designed to capture **non-linear patterns** while keeping the fitted function smooth and controlled.
It does this by:

- allowing small errors (inside an ε-tube)
- penalising larger errors
- using a **kernel function** (here: RBF) to bend the surface in non-linear ways

The result is a balanced, smooth 3D surface that follows the structure of the data without overfitting.

---

### Key parts of the printed model output

#### **SV type: eps-svr (regression)**
This is a regression-oriented variant of SVM.
Instead of predicting classes, it predicts a continuous outcome (Petal.Width).

#### **epsilon = 0.1**
Epsilon defines the width of the “tube” within which errors are ignored.

- larger ε → smoother surface
- smaller ε → more detailed, potentially wiggly fit

#### **cost C = 1**
C controls how strongly the model penalizes errors outside the ε-tube.

- high C → the model tries to fit almost everything
- low C → more regularized, more stable surface

Here, **C = 1** gives a well-balanced, conservative model.

#### **Gaussian RBF kernel**
The radial-basis kernel allows the model to bend and warp the fitted surface.
Without it (e.g., using a linear kernel), the result would be almost a flat plane.

RBF produces the smooth hills and valleys you later see in the 3D plot.

#### **sigma**
Sigma determines how wide the influence of each point is:

- small sigma → very wiggly surface
- large sigma → very smooth surface

Sigma is estimated automatically from the data.

#### **Number of support vectors**
Support vectors are the points that lie *outside* the ε-tube.
These are the data points that actually shape the regression surface.

In this example: **116 support vectors out of 150 flowers**
→ the data have quite a bit of non-linear structure.

#### **Training error**
Indicates how closely the model matches the training data.
A value around 0.12 means the average error is about 0.12 cm of Petal.Width — very reasonable.

---

### Summary

This step fits a **non-linear SVR model** that can describe how Sepal.Width and Sepal.Length jointly predict Petal.Width.
The RBF kernel gives the model flexibility, while parameters like ε and C keep the surface stable and smooth.
The trained model is then used in the next steps to generate predictions on a grid and to construct the interactive 3D surface.

```{r}
model <- svm_rbf(cost = 1.0) %>%
  set_engine("kernlab") %>%
  set_mode("regression") %>%
  fit(Petal.Width ~ Sepal.Width + Sepal.Length, data = iris)

  model
```

## Step 3 — Creating the prediction grid

To draw a smooth 3D surface, the model needs to be evaluated not only on the observed data points, but on a **dense grid** that covers the entire Sepal.Width × Sepal.Length space.

We therefore:

- compute the minimum and maximum of both predictors
- generate sequences (`xrange`, `yrange`) with a small step (`mesh_size = 0.02`)
- combine them into a 2D lattice using `meshgrid()`

This grid contains **thousands of (x, y) points**, evenly spaced across the whole plane.
These points will later receive predicted Petal.Width values, forming the smooth surface.

The grid itself has no biology in it — it is simply a mathematical canvas on which we paint the regression model.

```{r}
mesh_size <- .02
margin <- 0

x_min <- min(X$Sepal.Width) - margin
x_max <- max(X$Sepal.Width) - margin
y_min <- min(X$Sepal.Length) - margin
y_max <- max(X$Sepal.Length) - margin

xrange <- seq(x_min, x_max, mesh_size)
yrange <- seq(y_min, y_max, mesh_size)
str(yrange)

xy <- meshgrid(x = xrange, y = yrange)
xx <- xy$X
yy <- xy$Y
```

## Step 4 — Predicting values on the grid

The SVR model was trained on the real iris flowers, but here we need predictions for every point in the artificial grid.

We therefore:

1. reshape the grid matrices (`xx`, `yy`) into long column vectors
2. combine them into a two-column matrix (`final`)
3. feed all points into the trained model at once
4. reshape the predictions back into a matrix with the same dimensions as the grid

This gives us a predicted Petal.Width value for every location in the grid, turning the learned SVR function into a **smooth 3D surface**.

The result (`pred`) is a matrix perfectly aligned with the grid and ready to be plotted using plotly’s `add_surface()`.

```{r}
dim_val <- dim(xx)
dim_val

xx1 <- matrix(xx, length(xx), 1)
yy1 <- matrix(yy, length(yy), 1)
final <- cbind(xx1, yy1)

pred <- model %>%
  predict(final)

pred <- pred$.pred
str(pred)

z_pred <- matrix(pred, dim_val[1], dim_val[2])
str(z_pred)
```

## Step 5 — Construct interactive 3D plot

### Step 5a — Construct the basic 3D scatter plot

This first version shows only the **raw data** in three dimensions.
Before fitting any model, we visualise the actual measurements and the geometric shape of the data cloud.

```{r}

plot_ly(
  iris,
  x = ~Sepal.Width,
  y = ~Sepal.Length,
  z = ~Petal.Width
)


```

### Step 5b — Add observed points with marker styling

Now we add proper **3D markers**, giving the point cloud more structure.
This version will serve as the visual base on which the regression surface will be placed.

```{r}
plot_ly(
  iris,
  x = ~Sepal.Width,
  y = ~Sepal.Length,
  z = ~Petal.Width
) %>%
  add_markers(size = 4, name = "Observed flowers")

```

### Step 5c — Add the SVR regression surface

We now overlay the **predicted 3D surface** produced by the Support Vector Regression model.
The surface shows how the fitted function behaves across the entire grid of sepal measurements.

```{r}
plot_ly(
  iris,
  x = ~Sepal.Width,
  y = ~Sepal.Length,
  z = ~Petal.Width
) %>%
  add_markers(size = 4, name = "Observed flowers") %>%
  add_surface(
    x = xrange,
    y = yrange,
    z = z_pred,
    opacity = 0.7,
    name = "SVR prediction surface",
    showscale = TRUE
  )

```

### Step 5d — Adjust 3D scene layout

The final version improves readability:
custom axis titles, balanced camera angle, and a clean 3D scene.

```{r}
  plot_ly(
  iris,
  x = ~Sepal.Width,
  y = ~Sepal.Length,
  z = ~Petal.Width
) %>%
  add_markers(size = 4, name = "Observed flowers") %>%
  add_surface(
    x = xrange,
    y = yrange,
    z = z_pred,
    opacity = 0.7,
    name = "SVR prediction surface",
    showscale = TRUE
  ) %>%
  layout(
    scene = list(
      xaxis = list(title = "Sepal width"),
      yaxis = list(title = "Sepal length"),
      zaxis = list(title = "Petal width")
    )
  )


```



## Interacting with the 3D plot

The plotly figure is not just a static picture – you can explore the fitted surface interactively:

- **Rotate** – click and drag with the left mouse button to change the viewing angle.
- **Zoom** – use the mouse wheel or the *Zoom* tool in the toolbar to move closer or farther away.
- **Pan** – drag with the right mouse button (or with a two-finger drag on a trackpad) to shift the whole scene.
- **Reset view** – double-click inside the plot or click the *home* icon in the toolbar to go back to the default view.
- **Hover tooltips** – move the mouse over points to see the exact `Sepal.Width`, `Sepal.Length`, and `Petal.Width` values.
- **Toggle traces** – if the plot contains multiple layers, clicking entries in the legend turns individual traces on and off.
- **Download as image** – the camera icon saves the current view as a PNG file.
  (This works most reliably in a web browser; in some embedded viewers it may be limited.)

Some toolbar icons are shared between 2D and 3D plots, so a few tools may appear less useful here (for example, 2D lasso selection on a 3D surface). For this example, rotation, zooming, panning and hover tooltips are the most important interactions.