---
title: "3D regression surface: iris SVR example"
format: html
---

This page demonstrates how to build a **3D regression surface** using:

- two predictors: `Sepal.Width`, `Sepal.Length`
- one outcome: `Petal.Width`
- a flexible **support vector regression (SVR)** model
- interactive **plotly** surface rendering

The goal is to visualise how a smooth non–linear model passes through the cloud of observed flowers.

## Load data and packages

```{r}
library(tidyverse)
library(tidymodels)
library(kernlab)
library(pracma)
library(plotly)

iris_tbl <- as_tibble(iris)
glimpse(iris_tbl)
```

## Step 1 — Select predictors and outcome

```{r}
X <- iris_tbl %>%
  select(Sepal.Width, Sepal.Length)

y <- iris_tbl %>%
  select(Petal.Width)
```

## Step 2 — Fit SVR model

```{r}
model <- svm_rbf(cost = 1.0) %>%
  set_engine("kernlab") %>%
  set_mode("regression") %>%
  fit(Petal.Width ~ Sepal.Width + Sepal.Length, data = iris_tbl)
```

## Step 3 — Create prediction grid

```{r}
mesh_size <- .02
margin <- 0

x_min <- min(X$Sepal.Width) - margin
x_max <- max(X$Sepal.Width) - margin
y_min <- min(X$Sepal.Length) - margin
y_max <- max(X$Sepal.Length) - margin

xrange <- seq(x_min, x_max, mesh_size)
yrange <- seq(y_min, y_max, mesh_size)

xy <- meshgrid(x = xrange, y = yrange)
xx <- xy$X
yy <- xy$Y
```

## Step 4 — Predict values on grid

```{r}
dim_val <- dim(xx)

xx1 <- matrix(xx, length(xx), 1)
yy1 <- matrix(yy, length(yy), 1)
final <- cbind(xx1, yy1)

pred <- model %>%
  predict(final)

pred <- pred$.pred
z_pred <- matrix(pred, dim_val[1], dim_val[2])
```

## Step 5 — Construct interactive 3D plot

```{r}
fig <- plot_ly(
  iris_tbl,
  x = ~Sepal.Width,
  y = ~Sepal.Length,
  z = ~Petal.Width
) %>%
  add_markers(size = 4, name = "Observed flowers") %>%
  add_surface(
    x = xrange,
    y = yrange,
    z = z_pred,
    opacity = 0.7,
    name = "SVR prediction surface",
    showscale = TRUE
  ) %>%
  layout(
    scene = list(
      xaxis = list(title = "Sepal width"),
      yaxis = list(title = "Sepal length"),
      zaxis = list(title = "Petal width")
    )
  )

fig
```