---
title: "Two-layer scatterplot"
---

This task trains the basic ggplot workflow: mapping aesthetics, adding layers,
and reading a smooth trend.

## Goal
Replicate a scatterplot with a smooth trend line.  
Focus first on *structure* (layers), then on *style*.

## Data
We will use the built-in `cars` dataset (speed vs. stopping distance).

```{r setup}
# Load packages (no output)
library(tidyverse)
```

## Step 0: Preview the data

```{r step0}
cars
```

## Step 1: Start with a basic scatterplot

```{r step1}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point()
```

What to notice:
- `ggplot(data, aes(...))` defines *what* is plotted.
- `geom_point()` draws the points.

## Step 2: Improve readability of points

```{r step2}
ggplot(cars, aes(speed, dist)) +
  geom_point(size = 2)
```

Hints:
- `size` changes point size.
- `alpha` makes points semi-transparent (helps with overplotting).

## Step 3: Add a smooth trend line

```{r step3}
ggplot(cars, aes(speed, dist)) +
  geom_point(size = 2) +
  geom_smooth(se = TRUE)
```

Hints:
- `geom_smooth()` adds a model-based trend.
- `se = TRUE` shows a confidence band (gray ribbon).

## Step 4: Tune the smooth (optional but closer to target)

```{r step4}
ggplot(cars, aes(speed, dist)) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = TRUE)
```

Notes:
- LOESS gives a flexible curve (good for small datasets).
- Line color and thickness are styling details.

## Step 5: Final polish (labels + clean theme)

Hints:
- make points a little more transparent (70%)
- add labs and title
  - x-axis label "Speed (mph)"
  - y-axis label "stopping distance (ft)"
  - plot title "Cars: speed and stopping distance"
- set size of all text in the plot to 14 points


```{r step5}
ggplot(cars, aes(speed, dist)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    x = "Speed (mph)",
    y = "Stopping distance (ft)",
    title = "Cars: speed and stopping distance"
  ) +
  theme_minimal(base_size = 14)
```

---


## Extra challenges
1. Remove the gray confidence band.
2. Try a linear trend instead of LOESS.
3. Change the point color and add a matching smooth line.
