---
title: "Theme showcase: one plot, many looks"
code-fold: false
---

This page shows how different **ggplot2 themes** change the same plot.
We keep the data and geoms fixed, and only swap the theme.  
Use it as inspiration when polishing your own figures.

## Data

We use a custom-prepared subset of personality questionnaire data based on the
**Big Five Inventory (BFI)** included in the *psych* R package by
**William Revelle (Northwestern University)**.

Original data source:
Revelle, W. (2024). *psych: Procedures for Psychological, Psychometric, and Personality Research*.
Available from CRAN: https://cran.r-project.org/package=psych

What we use here:

- the original BFI dataset (25 self-report items + demographics),
- scored into five trait scales using `psych::scoreItems()`,
- cleaned, labeled, and saved as `bfi_traits.csv` for teaching purposes.

If you want to reproduce the processed file, see `src/data_prep/bfi_traits.R`.

<p>
  <a href="../assets/data/bfi_traits.csv" download>
    Download the dataset (bfi_traits.csv)
  </a>
</p>

```{r setup}
library(tidyverse)

# Load prepared dataset (created by src/data_prep/bfi_traits.R)
scores <- read_csv2("../assets/data/bfi_traits.csv")

# Base plot: keep geometry constant, no explicit theme here
base_plot <- scores %>%
  pivot_longer(
    cols = agree:openness,
    names_to = "trait",
    values_to = "score"
  ) %>%
  mutate(
    gender = factor(gender, labels = c("Male", "Female")),
    trait = factor(
      trait,
      levels = c("agree", "conscientious", "extraversion", "neuroticism", "openness"),
      labels = c("Agreeable", "Conscientious", "Extraversion", "Neuroticism", "Openness")
    )
  ) %>%
  ggplot(aes(x = trait, y = score, fill = gender)) +
  geom_boxplot() +
  labs(
    x = "BFI trait",
    y = "Score",
    title = "BFI scores by gender"
  ) +
  scale_fill_manual(
    name = "Gender",
    values = c("#008000", "#800080")
  ) +
  facet_grid(rows = vars(gender), labeller = label_value)
```

## Default look (no theme added)

This is what ggplot uses by default (`theme_gray()`).

```{r theme-default}
base_plot
```

## theme_gray()

The default ggplot theme with a light gray background.  
Good when you want a “standard R look” or to emphasize the data over styling.

```{r theme-gray}
base_plot + theme_gray()
```

## theme_minimal()

Very light, low‑ink design.  
Nice for slides and papers when you want the plot to feel airy and modern.

```{r theme-minimal}
base_plot + theme_minimal()
```

## theme_classic()

White background, clear axes, no gridlines.  
Great for simple teaching plots and for publications where grids feel cluttered.

```{r theme-classic}
base_plot + theme_classic()
```

## theme_bw()

Black‑and‑white theme with grids.  
A safe “journal‑style” default, especially for printing.

```{r theme-bw}
base_plot + theme_bw()
```

## theme_light()

Similar to `theme_gray()` but lighter and cleaner.  
Good when you want grids, but less visual weight.

```{r theme-light}
base_plot + theme_light()
```

## theme_linedraw()

Crisp black lines with minimal distractions.  
Useful for technical figures and when you expect black‑and‑white reproduction.

```{r theme-linedraw}
base_plot + theme_linedraw()
```

## theme_dark()

Dark background with light grids.  
Can work on dark slides, or when you want a dramatic contrast.

```{r theme-dark}
base_plot + theme_dark()
```

## theme_void()

Removes everything except the data.  
Good for artistic visuals, insets, or when you add your own annotations later.

```{r theme-void}
base_plot + theme_void()
```


