---
title: "Tufte’s principles (and how to avoid lying with axes)"
code-fold: false
---

This page is a small, visual reminder of a few ideas from **Edward Tufte**:

- Graphics should maximize *data-ink* (ink that shows data) and minimize *non‑data ink*.
- The goal is to show the *truthful* size of effects, not to “win” with a dramatic chart.
- A common way to mislead is to truncate an axis.

We’ll illustrate that with a simple example from the `gapminder` dataset.

```{r setup}
library(tidyverse)
library(gapminder)
```

## Example: life expectancy in Peru before 1980

```{r data}
life_exp_pre80 <- gapminder %>%
  filter(country == "Peru", year < 1980)
```

### 1) A misleading plot (truncated y‑axis)

This plot *zooms* the y‑axis to a narrow range.  
It makes the improvement in life expectancy look **much larger** than it really is.

```{r misleading}
life_exp_pre80 %>%
  ggplot(aes(x = year, y = lifeExp, fill = as.factor(year))) +
  geom_col() +
  theme_light() +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(43.9, 60)) +
  geom_hline(yintercept = 43.9, color = "red", size = 2) +
  geom_hline(yintercept = 58.45, color = "red", size = 2) +
  geom_label(
    aes(x = 1960, y = 58.45, fill = NULL),
    label = "Ratio is about 4.5",
    size = 10
  )
```

### 2) The honest version (axis starts at zero)

Same data, same geom, same colors.  
Only the axis range is fixed so the magnitude is not exaggerated.

```{r honest}
life_exp_pre80 %>%
  ggplot(aes(x = year, y = lifeExp, fill = as.factor(year))) +
  geom_col() +
  theme_light() +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 60)) +
  geom_hline(yintercept = 43.9, color = "red", size = 2) +
  geom_hline(yintercept = 58.45, color = "red", size = 2) +
  geom_label(
    aes(x = 1960, y = 58.45, fill = NULL),
    label = "Ratio is about 1.3",
    size = 10
  )
```

### Quick check (the real ratio)

```{r ratio}
max_life_exp <- life_exp_pre80 %>% slice_max(lifeExp)
min_life_exp <- life_exp_pre80 %>% slice_min(lifeExp)

max_life_exp$lifeExp / min_life_exp$lifeExp
```

### 3) A deliberately “bad” example (what not to do)

This plot is purposely overloaded.
It demonstrates the classic mistakes that Tufte warns about:
too much background, heavy gridlines, loud colors, and no clear message.

Goal: Help students see how easily a plot can become cluttered.

```{r crazy}
gapminder %>%
    filter(country == "Peru", year < 1980) %>%
    ggplot(aes(x = year, y = lifeExp, fill = as.factor(year))) +
        geom_col() +
        theme(
            panel.grid.major = element_line(size = 1, linetype = 'dashed', colour = "green"),
            panel.grid.minor = element_line(size = 1, linetype = 'dashed', colour = "blue"),
            panel.background = element_rect(fill = "red", colour = "blue"),
            plot.background = element_rect(fill = "green")
        )
```


### 4) A cleaned-up reference version

Same data, but now the plot is readable.
The layout is simple, the background is neutral, and unnecessary visual elements are removed.

Goal: Show the transition from visual noise to a reasonable starting point.

```{r crazy-fixed}
gapminder %>%
    filter(country == "Peru", year < 1980) %>%
    ggplot(aes(x = year, y = lifeExp, fill = as.factor(year))) +
        geom_col() +
        theme_light() +
        theme(legend.position = "none")
```

### 5) Highlighting one meaningful value (Tufte-ish)

Here we finally work with visual hierarchy:
one bar (1967) is emphasized while everything else fades into the background.
No gridlines, minimal text, and clear focus on the information that matters.

Goal: Demonstrate that a minimalist plot can be far more informative than a decorative one.


```{r highlight}

gapminder %>%
    filter(country == "Peru", year < 1980) %>%
    ggplot(aes(x = year, y = lifeExp, fill = (year == 1967))) +
        geom_col() +
        theme_minimal() +
        theme(
            axis.text.y = element_blank(),
            legend.position = "none") +
    scale_fill_manual(values = c("gray", "red")) +
    geom_text(
    aes(x = year, y = lifeExp, label = round(lifeExp)), nudge_y = 0.25, size = 7, vjust = 2, color = "black")

```

## Take-home message

- If your axis does not start at zero, explain why and check that it does not distort the story.
- Prefer clarity over decoration.
- The “best” theme is the one that makes the data the easiest to read.
- Highlight only what carries meaning. Silence around the important signal helps it stand out.
